---
title: 'DATA 621: Blog 2'
author: "Andrew Bowen"
date: "2023-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)
```


We can use the built-in `iris` dataset, which is a good sampel dataset for classification tasks. In this case, we'll want to predict the multiclass `Species` variable
```{r}
data("iris")
head(iris)
```


```{r}
# Load built-in iris dataset
data("iris")
```


Let's plot counts of our `Species` categorical variable to get a sense of our class sizes. It looks like we have as balanced of a dataset as we could hope for over this variable.
```{r}
ggplot(data=iris, aes(x=Species)) + geom_bar()
```

## Data Wrangling
```{r species-as-factor}
# Re-write species as a factor
iris$species <- as.factor(iris$Species)
```

```{R train-test}
# Breaking out training and test datasets
split <- initial_split(iris, prop = 0.8, strata = species)
train <- split %>% 
         training()
test <- split %>% 
        testing()
```


## Modeling
Let's use the [`multinom_reg` model](https://parsnip.tidymodels.org/reference/multinom_reg.html) that comes from `parsnips`. This can allow us to predict multi-class variables based on input variables. In this case, we'll use `Sepal.Length` and `Petal.Width` to build our classifier.
```{r}
model <- multinom_reg() %>% fit(species ~ Sepal.Length + Petal.Width, train)
```

```{r}
tidy(model)
```

## Model Evaluation
```{R}
# Predict class on test data
pred_class <- predict(model,
                      new_data = test,
                      type = "class")
# predict relative class probabilities
pred_prob <- predict(model,
                      new_data = test,
                      type = "prob")
```


```{R}
results <- test %>%
           select(species) %>%
           bind_cols(pred_class, pred_prob)

results$match <- ifelse(results$.pred_class == results$species, TRUE, FALSE)
```


```{r}
ggplot(results, aes(x=match)) + geom_bar() + labs(x="Prediction Matches Actual Class", y="Count of Flowers", title="Multinomial Regression Results, Iris Dataset")
```



```{r}
(nrow(results %>% filter(match==TRUE)) / nrow(results) )
```

We see an accuracy of 100%, which we should take with a grain of salt given the small size of our dataset





