---
title: 'DATA621: Blog 1'
author: "Andrew Bowen"
date: "2023-10-22"
output: html_document
---
## Power Transforms in R

The R-function `boxCox` from the `car` package can be used to implement maximum likelihood transformations of data when modeling. This can be used to transform data to create a more linear relationship between predictor and dependent variables.
```{r message=FALSE, echo=FALSE}
library(tidyverse)
library(car)
library(MASS)
```


We'll use the built-in `mtcars` dataset. Let's plot our horsepower variable against our mpg, which is the variable we'd like to predict:
```{r}
data(mtcars)
plot(mtcars$hp, mtcars$mpg)
```

This looks like it *could* be a linear relationship between horsepower and miles per gallon, but we can build a simple linear model in order to evaluate if this is a simple OLS regression:
```{r}
# Create model to predict MPG from Horsepower
lm_hp <- lm(mpg ~ hp, mtcars)
summary(lm_hp)
```

We can plot our model to begin evaluating how effective it is at predicting MPG. The `plot` function called on an `lm` object is a useful tool here.

```{r}
plot(lm_hp)
```


Looking at the above plots, there's a bit of a pattern in our residuals vs fitted plot, and some tail behavior in our QQ plot indicating a transform could be used here. We can implement a [Box-Cox transformation](https://builtin.com/data-science/box-cox-transformation-target-variable#) to improve our fit.



```{r}
bc <- boxCox(lm_hp)
```

Our box-cox transform plot gives us a non-zero $\lambda$ parameter value, around 0.05, indicating we're [close to a log-transformation](https://www.r-bloggers.com/2022/10/box-cox-transformation-in-r/)

We can see similar results below by applying a simple log transformation:
```{r}
lm_log_hp <- lm(mpg ~ log(hp), mtcars)
summary(lm_log_hp)
```

And plot the results of our log-transform model

```{r}
plot(lm_log_hp)
```

